Colors and residual-based Shadings in Strucplots

### 2005-08-24 trunk/vcd/inst/doc/shading.Rnw
viewport names changed to "obj:dim1=level1,...,dimn=leveln" syntax.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@444 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine=R,eps=TRUE}
\setkeys{Gin}{width=0.8\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
@

@@

As an example, consider the Admissions to the University of California
in Berkely (UCB) data. In the table collapsed over departments, we
would like to highlight the (incidentally wrong) impression that there
were to much male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

<<ucb,fig=FALSE,echo=TRUE,height=6,width=7,results=verbatim>>=
ucb <- margin.table(UCBAdmissions, 1:2)
ucb

fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2)
fill_colors

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<ucbfig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(ucb, gp = gpar(fill = fill_colors, col = 0))
@
\caption{Mosaic plot for the UCBAdmissions data with highlighted cells.}
\label{fig:ucb}
\end{center}
\end{figure}

@@

If the parameters specified in the \code{gpar} object are incomplete,
strucplot functions will recycle them along the last splitting
dimension. In the following example based on the Survival on the
Titanic data, we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

\begin{Sinput}
> mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<recycling,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
@
\caption{Recycling of parameters.}
\label{fig:recycling}
\end{center}
\end{figure}

@@

\begin{Sinput}
> obs <- margin.table(UCBAdmissions, 1:2)
> exp <- independence_table(obs)
> res <- (obs - exp) / sqrt(exp)
> res
          Gender
Admit           Male    Female
  Admitted  4.784093 -5.793466
  Rejected -3.807325  4.610614

> gpobj <- shading_binary()(res)
> gpobj
$fill
     [,1]      [,2]     
[1,] "#96A2DF" "#E18E9E"
[2,] "#E18E9E" "#96A2DF"

> mosaic(obs, gp = gpobj)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<binary,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
obs <- margin.table(UCBAdmissions, 1:2)
exp <- independence_table(obs)
res <- (obs - exp) / sqrt(exp)
gpobj <- shading_binary()(res)
mosaic(obs, gp = gpobj)
@
\caption{Binary shading.}
\label{fig:binary}
\end{center}
\end{figure}

@@

\begin{Sinput}
> mosaic(obs, gp = shading_binary)
\end{Sinput}

@@

As an example, we will visualize the (in)dependence of hair and eye
color in the HairEyeColor data set (see Figure \ref{fig:haireye})

\begin{Sinput}
> haireye <- margin.table(HairEyeColor, 1:2)
> mosaic(haireye, shade = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<haireye,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, shade = TRUE)
@
\caption{Shaded residuals in the HairEyeColor data set---two cut-off points}
\label{fig:haireye}
\end{center}
\end{figure}

@@

\begin{Sinput}
> mosaic(haireye, shade = TRUE, 
         gp_args = list(interpolate = 1:4, h = c(100, 200)))
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<interpolate,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(haireye, shade = TRUE, 
       gp_args = list(interpolate = 1:4, h = c(100, 200)))
@
\caption{Shaded residuals in the HairEyeColor data set---four cut-off points}
\label{fig:interpolate}
\end{center}
\end{figure}

\noindent Note that we also changed the hue values for purely cosmetic
reasons. The third remaining dimension, the luminance, is used for
visualizing the significance of a test statistic. The $p$ value can
be supplied to the \code{p.value} argument. If none is specified, it
is computed from a Chi-squared distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
otherwise dark colors are employed. These colors are computed using
the two luminance values supplied to the \code{l} argument. The
following example shows the relationship of home goals and away
goals of the German Bundesliga in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
Chi-squared test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

\begin{Sinput}
> bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
> mosaic(bl, shade = TRUE)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<bundesliga,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, shade = TRUE)
@
\caption{Non-significant Chi-squared test using part of the Bundesliga data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

@@

\begin{Sinput}
> art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
> mosaic(art, shade = TRUE, gp = shading_max)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
mosaic(art, gp = shading_max)
@
\caption{Significant maximum test on female patients of the Arthritis
  data.}
\label{fig:maximum}
\end{center}
\end{figure}

@@

\begin{Sinput}
> mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
@
\caption{The Bundesliga data for 1995 using the Friendly shading.}
\label{fig:friendly}
\end{center}
\end{figure}

@@

\begin{Sinput}
shading_continuous <- function(observed = NULL, residuals = NULL,
                               expected = NULL, df = NULL, n = 5000, ...)
{
  ## function works only for two-way tables
  stopifnot(length(dim(observed)) == 2)

@@

mosaic(haireye, gp = shading_continuous)
\end{Sinput}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[h]
\begin{center}
<<continuous,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
shading_continuous <- function(observed = NULL, residuals = NULL,
                               expected = NULL, df = NULL, n = 5000, ...)
{
  ## function works only for two-way tables
  stopifnot(length(dim(observed)) == 2)

mosaic(haireye, gp = shading_continuous)
@
\caption{The HairEyeColor data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}

>>>>

\SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
\setkeys{Gin}{width=0.7\textwidth}

<<preliminaries,echo=FALSE,results=hide>>=
options(width=90)
set.seed(1071)
library(vcd)
data(Titanic)
@

@@

As an example, consider the `Admissions to the University of California
in Berkely (UCB)' data. In the table collapsed over departments, we
would like to highlight the (incidentally wrong) impression that there
were to much male students accepted compared to the presumably
discriminated female students (see Figure \ref{fig:ucb}):

<<ucb,results=verbatim>>=
(ucb <- margin.table(UCBAdmissions, 1:2))

(fill_colors <- matrix(c("dark cyan","gray","gray","dark magenta"), ncol = 2))

@@

\begin{figure}[h]
\begin{center}
<<ucbfig,fig=TRUE,echo=FALSE>>=
<<ucb>>
@
\caption{Mosaic plot for the UCBAdmissions data with highlighted cells.}
\label{fig:ucb}
\end{center}
\end{figure}

@@

If the parameters specified in the \code{gpar} object are incomplete,
strucplot functions will recycle them along the last splitting
dimension. In the following example based on the `Survival on the
Titanic' data, we will highlight all cells corresponding to survived
passengers (see Figure \ref{fig:recycling}):

<<recycling>>=
mosaic(Titanic, gp = gpar(fill = c("gray","dark magenta")))
@

\begin{figure}[h]
\begin{center}
<<recyclingfig,fig=TRUE,echo=FALSE>>=
<<recycling>>
@
\caption{Recycling of parameters.}
\label{fig:recycling}
\end{center}
\end{figure}

@@

<<binary,results=verbatim>>=
obs <- margin.table(UCBAdmissions, 1:2)
exp <- independence_table(obs)
(res <- (obs - exp) / sqrt(exp))

(gpobj <- shading_binary()(res))

mosaic(obs, gp = gpobj)
@

\begin{figure}[h]
\begin{center}
<<binaryfig,fig=TRUE,echo=FALSE>>=
<<binary>>
@
\caption{Binary shading.}
\label{fig:binary}
\end{center}
\end{figure}

@@

<<binaryfun>>=
mosaic(obs, gp = shading_binary)
@

@@

As an example, we will visualize the (in)dependence of hair and eye
color in the `Hair and Eye Color' data set (see Figure \ref{fig:haireye})

<<haireye>>=
haireye <- margin.table(HairEyeColor, 1:2)
mosaic(haireye, shade = TRUE)
@

\begin{figure}[h]
\begin{center}
<<haireyefig,fig=TRUE,echo=FALSE>>=
<<haireye>>
@
\caption{Shaded residuals in the HairEyeColor data set---two cut-off points}
\label{fig:haireye}
\end{center}
\end{figure}

@@

<<interpolate>>=
mosaic(haireye, shade = TRUE, 
       gp_args = list(interpolate = 1:4, h = c(100, 200)))
@

\begin{figure}[h]
\begin{center}
<<interpolatefig,fig=TRUE,echo=FALSE>>=
<<interpolate>>
@
\caption{Shaded residuals in the HairEyeColor data set---four cut-off points}
\label{fig:interpolate}
\end{center}
\end{figure}

\noindent Note that we also changed the hue values for purely cosmetic
reasons. The third remaining dimension, the luminance, is used for
visualizing the significance of a test statistic. The $p$ value can
be supplied to the \code{p.value} argument. If none is specified, it
is computed from a Chi-squared distribution with \code{df} degrees of
freedom. The \code{level} argument is used to specify the confidence
level:  if \code{p.value} is smaller than \code{1 - level}, bright colors are used, 
otherwise dark colors are employed. These colors are computed using
the two luminance values supplied to the \code{l} argument. The
following example using the `Bundesliga' data 
shows the relationship of home goals and away
goals of the German football league in 1995: although there are two
``larger'' residuals (one greater than 2, one less then $-2$), the
Chi-squared test does not reject the null hypothesis of
independence. Consequently, the colors appear dark (see Figure \ref{fig:bundesliga}):

<<bundesliga>>=
bl <- xtabs(~ HomeGoals + AwayGoals, data = Bundesliga, subset = Year == 1995)
mosaic(bl, shade = TRUE)
@

\begin{figure}[h]
\begin{center}
<<bundesligafig,fig=TRUE,echo=FALSE>>=
<<bundesliga>>
@
\caption{Non-significant Chi-squared test using part of the Bundesliga data.}
\label{fig:bundesliga}
\end{center}
\end{figure}

@@

<<arthritis>>=
art <- xtabs(~ Improved + Treatment, data = Arthritis, subset = Sex == "Female")
mosaic(art, gp = shading_max)
@

\begin{figure}[h]
\begin{center}
<<arthritisfig,fig=TRUE,echo=FALSE>>=
<<arthritis>>
@
\caption{Significant maximum test on female patients of the Arthritis
  data.}
\label{fig:maximum}
\end{center}
\end{figure}

@@

<<friendly>>=
mosaic(bl, gp = shading_Friendly, legend = legend_fixed, zero_size = 0)
@

\begin{figure}[h]
\begin{center}
<<friendlyfig,fig=TRUE,echo=FALSE>>=
<<friendly>>
@
\caption{The Bundesliga data for 1995 using the Friendly shading.}
\label{fig:friendly}
\end{center}
\end{figure}

@@

<<continuous>>=
shading_continuous <- function(observed = NULL, residuals = NULL,
                               expected = NULL, df = NULL, n = 5000, ...)
{
  ## function works only for two-way tables
  stopifnot(length(dim(observed)) == 2)

@@

mosaic(haireye, gp = shading_continuous)
@

\begin{figure}[h]
\begin{center}
<<continuousfig,fig=TRUE,echo=FALSE>>=
<<continuous>>
@
\caption{The HairEyeColor data with continuous shading.}
\label{fig:continuous}
\end{center}
\end{figure}
